\documentclass[main.tex]{subfiles}

\begin{document}
    \chapter*{Abstract}
    Neutron stars (NSs) are compact remnants of massive stars formed in supernova explosions, with properties intricately linked to fundamental physical laws, the evolution of their progenitors, and the characteristics of the explosion. This thesis investigates the structure and evolution of neutron stars, beginning with an overview of stellar astrophysics and the processes leading to compact object formation. Special emphasis is placed on the complex internal structure of neutron stars, particularly the challenges posed by the extreme conditions found within their cores. It then presents two separate research papers in subsequent chapters,

    Through detailed numerical modeling, the first research paper examines neutron stars formed from the collapse of intermediate-mass stars through electron-capture supernovae (ECSN). Stars with initial masses of $7–11\msun$ that lose their hydrogen envelopes may experience outcomes ranging from ECSN to thermonuclear supernovae, depending on their final core composition and structure. Modeling of stripped, helium stars reveals that cores around $1.35–1.37\msun$ can initiate explosive oxygen burning at relatively low densities, leading to thermonuclear explosions if residual carbon exceeds $\sim 0.005\msun$, while lower carbon content favors ECSN. The study further finds that (C)ONe Type Ia supernovae are more likely at higher metallicities, with late-stage expansion possibly triggering binary interactions in both progenitor types.

    The second research paper explores the formation of twin compact stars through rapid phase transitions in NS cores due to mass accretion in low-mass X-ray binaries (LMXBs).
    Neutron stars in binary systems can experience episodes of mass accretion from their companion stars. During such mass transfer episodes, the neutron star’s mass---and therefore its central density---gradually increases. As the core density rises, it may reach a threshold where a phase transition from hadronic matter to deconfined quark matter can occur. This phase transition marks a fundamental shift from a purely hadronic composition, which is dominated by neutrons, protons, and electrons, to a more exotic state where quarks are no longer confined within individual nucleons, yielding a quark core. Simulations reveal that phase transitions may occur during the LMXB phase in compact binaries or during spin-down in wider systems, possibly leading to eccentric binary millisecond pulsars (MSPs). If these transitions involve secondary kicks exceeding 20 km/s, they may disrupt the binary, forming isolated MSPs or reconfiguring them to ultra-wide orbits. The eccentricity distribution of binary MSPs could provide constraints on phase transitions in dense nuclear matter.

    The thesis concludes with a comprehensive summary of the key findings and insights derived from the research conducted. In this final section, the main conclusions drawn from the study are highlighted, emphasizing their significance in the broader context of the field. Additionally, a discussion is provided on the potential implications and ramifications of this work, including its impact on future research directions, practical applications, and theoretical advancements. The conclusion also reflects on any limitations of the current study and suggests areas for further investigation, thereby offering a holistic view of the work's contribution and its relevance to ongoing scientific inquiry.



    

    % Using novel numerical modeling, this work examines neutron stars formed from the collapse of intermediate-mass stars through electron-capture supernovae (ECSN), revealing that some stars may instead undergo thermonuclear runaway, preventing NS formation. This analysis further constrains the conditions necessary for such outcomes and explores their implications for the observed distribution of NS masses.

    % Additionally, this thesis models the evolution of neutron stars within binary systems, exploring phase transitions induced by mass accretion that could result in hybrid stars with exotic matter compositions.
    
    % Neutron stars (NSs) are the compact remnants of massive stars formed in supernova explosions. Their properties depend sensitively on the properties of fundamental physical laws, the evolution of their progenitors, and the physical characteristics of the explosion. This thesis, [...]

    % Chapter~1 provides a brief and necessary introduction to stellar astrophysics and is divided into two parts; In the first part, I make an attempt to equip the reader with the essentials of stellar structure and evolution that leads to the formation of compact objects. In the second part, I focus on neutron stars where I examine their structure and the complexities arising with extreme dense matter found in their interior.
 

\end{document}