\documentclass[main.tex]{subfiles}

\begin{document}
\chapter{The Equation of State of Ideal Gases}\label{apx:eos}
    {\hypersetup{linkcolor=black, pdfborder={0 0 0}}
        \minitoc
    }
\section{Deriving the equation of state: A macroscopic approach}
In thermodynamics, the equation of state is a fundamental relationship that defines the state of a material system under specific physical conditions. Typically, it is expressed in the form $P=P(\rho,T)$, where $P$ denotes the pressure, $\rho$ the density, and $T$ the temperature. 

To deepen our understanding, it is often useful to express the equation of state in differential form.  This differential representation can be derived by initially expressing the equation of state as a generalized power law:
\begin{equation}
    P = \rho^{\chi_\rho} T^{\chi_T},
\end{equation}
where $\chi_\rho$ and $\chi_T$ are exponents that describe how pressure $P$, varies with changes in density $\rho$, and temperature $T$, respectively. The derivation of the differential form begins by applying the principles of partial differentiation to this power law, resulting in the total differential of pressure as:
\begin{equation*}
    dP = \left( \frac{\partial P}{\partial \rho} \right)_T d\rho + \left( \frac{\partial P}{\partial T} \right)_\rho dT.
\end{equation*}
Expanding this differential using the initial power law yields:
\begin{equation*}
    dP = \chi_\rho T^{\chi_T} \rho^{\chi_\rho - 1} d\rho +  \chi_T \rho^{\chi_\rho} T^{\chi_T - 1} dT,
\end{equation*}
which, upon further manipulation, can be expressed in terms of relative changes:
\begin{equation}
    \frac{dP}{P} = \chi_\rho \frac{d\rho}{\rho} + \chi_T \frac{dT}{T}.
\end{equation}
This equation succinctly captures the relative change in pressure in response to infinitesimal changes in density and temperature. 

By comparing the relations:
\begin{align*}
    \label{apx:eq:differential_form_of_eos}
    dP &= \left( \frac{\partial P}{\partial \rho} \right)_T d\rho + \left( \frac{\partial P}{\partial T} \right)_\rho dT, \\\\
    \frac{dP}{P} &= \chi_\rho \frac{d\rho}{\rho} + \chi_T \frac{dT}{T},
\end{align*}
we find that the exponents $\chi_\rho$ and $\chi_T$ are given by the equations:
\begin{align*}
    \chi_\rho &= \frac{\rho}{P} \left( \frac{\partial P}{\partial \rho} \right)_T, \\\\
    \chi_T &= \frac{T}{P} \left( \frac{\partial P}{\partial T} \right)_\rho.
\end{align*}

However, one can find an even more elegant expression for the aforementioned exponents as follows:
\begin{align*}
     P &= \rho^{\chi_\rho} T^{\chi_T} \Rightarrow \log \,P = \log \,\left( \rho^{\chi_\rho} T^{\chi_T} \right) = \log \,\rho^{\chi_\rho} + \log \,T^{\chi_T} \Rightarrow \\\\
     &\Rightarrow \log \,P = \chi_\rho \log \,\rho + \chi_T \log \,T.
\end{align*}
The total differential then is:
\begin{equation*}
    d\log \,P = \left( \frac{\partial \log \,P}{\partial \log \,\rho} \right)_T d \log \,\rho + \left( \frac{\partial \log \,P}{\partial \log \,T} \right)_\rho d \log \,T,
\end{equation*}
so that:
\begin{equation*}
    d \log \,P = \chi_\rho d\log \,\rho + \chi_T d \log \,T,
\end{equation*}
and by comparing the last two equations, it finally emerges that:
\begin{align}
    \chi_\rho &= \frac{\rho}{P} \left( \frac{\partial P}{\partial \rho} \right)_T = \left( \frac{\partial \log \,P}{\partial \log \,\rho} \right)_T, \label{apx:eq:chi_rho}\\\nonumber\\
    \chi_T &= \frac{T}{P} \left( \frac{\partial P}{\partial T} \right)_\rho = \left( \frac{\partial \log \,P}{\partial \log \,T} \right)_\rho. \label{apx:eq:chi_T}
\end{align}
This derivation elegantly illustrates how the exponents $\chi_\rho$ and $\chi_T$ can be expressed in terms of logarithmic derivatives. By applying a logarithmic transformation to the equation of state, we obtain a clear and concise formula for these exponents, linking them directly to the partial derivatives of the logarithm of pressure with respect to the logarithms of density and temperature. 
In the most general case, the exponents $\chi_\rho$ and $\chi_T$ may depend on $\rho$ and $T$. However, if they are approximately constant, the equation of state can be written as:
\begin{equation}
    \label{apx:eq:eos_chi_rho_chi_T}
    P = P_0 \,\rho^{\chi_\rho} \,T^{\chi_T}.
\end{equation}
This formulation provides a simplified yet powerful representation of the equation of state, assuming that the exponents $\chi_\rho$ and $\chi_T$ remain invariant over the range of interest. Here, $P_0$ serves as a reference pressure, encapsulating the initial conditions and acting as a normalization factor for the system.



\section{Deriving the equation of state: A microscopic approach}
In the previous section, we examined the derivation of the generalized equation of state, Eq.~\eqref{apx:eq:eos_chi_rho_chi_T}, which captures the macroscopic response of a gas to changes in density and temperature, parameterized by the exponents $\chi_\rho$ and $\chi_T$.
This formulation provides an elegant link between thermodynamic properties and the statistical behavior of gas molecules across a broad range of conditions.
In this section, we derive an alternative form of the generalized equation of state, which offers a complementary perspective by directly relating the pressure of a gas to its microscopic kinetic energy and particle density.

A key focus in studying the thermodynamic properties of an ideal gas is understanding the distribution of particle momenta within the gas, represented by $n(p)$. The distribution function $n(p)dp$ describes the number of particles per unit volume with momentum $p$ within the interval $p \in [p, p+dp]$. Given $n(p)$ important physical quantities---such as the number density, energy density, and pressure---can be determined from the following integrals:
\begin{enumerate}
    \item \textit{Number density}
        \begin{equation}\label{apx:eq:number_density_integral}
            n = \int_{0}^{\infty} n(p) dp,
        \end{equation}
        
    \item \textit{Energy Density}
        \begin{equation}\label{apx:eq:energy_density_integral}
            u = \int_{0}^{\infty} E_{\text{kin}} n(p) dp = n \langle E_{\text{kin}} \rangle,
        \end{equation}
        
    \item \textit{Pressure}
        \begin{equation}\label{apx:eq:pressure_integral} 
            P = \frac{1}{3}\int_{0}^{\infty} p \,v_p \,n(p) dp = \frac{1}{3} n \langle p \,v_p \rangle.
        \end{equation}
\end{enumerate}
Equation~\eqref{apx:eq:number_density_integral}  is straightforward, stemming directly from the definition of $n(p)$. It computes the total number of particles per unit volume by integrating the momentum distribution over all possible momenta.
The second relationship for energy density (Eq.~\eqref{apx:eq:energy_density_integral}) follows from the definition of an ideal gas, where the internal energy is purely kinetic. Here, $E_\mathrm{kin}$ represents the kinetic energy of a particle with momentum $p$ and velocity $v_p$. This equation yields the total kinetic energy per unit volume, effectively providing an average of the kinetic energy across the particle distribution. The equation for pressure, however, requires a more in-depth examination. Derived from the kinetic theory of gases, it is based on the principle that pressure results from particle collisions with the container walls, with each particle transferring momentum proportional to $p\,v_p$ per unit area per unit time. The factor $1/3$ accounts for the isotropic nature of particle motion in a gas, assuming momentum transfer is uniformly distributed across the three spatial dimensions, as described by the equipartition theorem. This integral captures each particle's contribution, with momentum $p$ and velocity $v_p$, to the total pressure exerted by the gas.

Having defined the quantities above, we now seek to establish a relationship between pressure and internal energy. According to special relativity, the momentum and velocity of particles are related to their energy by the following equations:
\begin{equation}
    \epsilon^2 = p^2 c^2 + m^2 c^4,
\end{equation}
and
\begin{align}
    \nonumber \frac{\partial}{\partial p} (\epsilon^2) &= \frac{\partial}{\partial p} (p^2 c^2 + m^2 c^4) \Rightarrow 2\epsilon \frac{\partial \epsilon}{\partial p} = 2p c^2 \Rightarrow \\\nonumber\\
    &\Rightarrow \frac{\partial \epsilon}{\partial p} \equiv v_p = \frac{pc^2}{\epsilon}.
\end{align}
Using these relations, we can derive connections between pressure and internal energy for an ideal gas in both the non-relativistic (NR) and extremely relativistic (ER) limits:
\begin{itemize}
    \item \textbf{NR limit}: In this scenario, momenta are much smaller than $mc$ ($p \ll mc$)

    $$\epsilon^2 = p^2 c^2 + m^2 c^4 \Rightarrow \epsilon^2 = c^2 \left(p^2 + m^2c^2 \right).$$
    Given $p \ll mc$, we approximate $\epsilon \simeq mc^2$, therefore:
    $$v_p = \frac{pc^2}{\epsilon} = \frac{p}{m}.$$
    The kinetic energy is given by $E_{\text{kin}} = p^2/2m$ so that $\langle p v_p \rangle = \langle \frac{p^2}{m} \rangle = 2 \langle E_{\text{kin}} \rangle $.
    Finally, from the integral relations \eqref{apx:eq:energy_density_integral} and \eqref{apx:eq:pressure_integral}, it follows that:
    \begin{equation}
        \label{apx:eq:pressure_internal_energy_relation_for_nr_case}
        P = \frac{2}{3}u.
    \end{equation}


    \item \textbf{ER limit}: In this case, we have $p \gg mc$. This means $\epsilon = pc$, hence $v_p = c \longrightarrow \langle p v_p \rangle = \langle pc \rangle = \langle \epsilon \rangle$.
    Consequently, from the integral relations \eqref{apx:eq:energy_density_integral} and \eqref{apx:eq:pressure_integral}, it follows that:
    \begin{equation}
        \label{apx:eq:pressure_internal_energy_relation_for_er_case}
        P = \frac{1}{3}u.
    \end{equation}
\end{itemize}
Generalizing, we can write:
\begin{equation}\label{apx:eq:second_generalized_eos}
    P = \zeta n \langle E \rangle,
\end{equation}
where $\zeta = 2/3$ or $\zeta = 1/3$, for the non-relativistic and relativistic limits, respectively. This aligns with our intuition that the pressure arising from particle collisions would be proportional to the particle density and their kinetic energy:
\begin{equation*}
    P \sim n mv^2 \sim n E_{\text{kin}}.
\end{equation*}


Though at first glance the equations \eqref{apx:eq:eos_chi_rho_chi_T} and \eqref{apx:eq:second_generalized_eos} may seem distinct, they are inherently connected through the principles of statistical mechanics. The former emphasizes the macroscopic response to changes in physical conditions, encoded in the exponents $\chi_\rho$ and $\chi_T$, while the latter explores the microscopic origins of pressure in terms of particle density and kinetic energy. To illustrate their connection we will examine their application in two ideal gas cases: the classical Maxwell-Boltzmann ideal gas and the ideal quantum Fermi gas.

\subsection{The classical ideal gas}
In an ideal gas, the momentum distribution is given by the Maxwell-Boltzmann distribution:
\begin{equation}
    n(p) dp = \frac{n}{(2\pi m k_BT)^{3/2}} \exp\left(- \frac{p^2}{2mk_BT}\right) 4\pi p^2 dp,
\end{equation}
therefore, according to the relationship derived earlier for the non-relativistic case (Eq.~\ref{apx:eq:pressure_internal_energy_relation_for_nr_case}), the equation of state for an ideal gas is given by:
\begin{equation}\label{apx:eq:pgas}
    P_{\text{gas}} = \frac{2}{3}u = \frac{2}{3} n \langle E_{\text{kin}} \rangle.
\end{equation}
To derive the expression for the mean kinetic energy, we begin with the Maxwell-Boltzmann distribution for the speeds of gas particles, which is given by:
\begin{equation*}
    f(v)dv = 4\pi \left( \frac{m}{2\pi k_BT} \right)^{3/2} v^2 \exp\left(-\frac{mv^2}{2k_BT}\right)dv.
\end{equation*}
The mean kinetic energy, $\langle E_{\text{kin}} \rangle$, of a particle in an ideal gas can be expressed by integrating over all possible speeds, weighted by the probability distribution:
\begin{equation*}
    \langle E_{\text{kin}} \rangle = \int_{0}^{\infty} \frac{1}{2} m v^2 f(v) dv.
\end{equation*}
Substituting the expression for $f(v)dv$ into the equation for $\langle E_{\text{kin}} \rangle$ yields:
\begin{equation*}
    \langle E_{\text{kin}} \rangle = \int_{0}^{\infty} \frac{1}{2} m v^2 4\pi \left( \frac{m}{2\pi k_BT} \right)^{3/2} v^2 \exp\left(-\frac{mv^2}{2k_BT}\right) dv.
\end{equation*}
This integral can be solved using integration by parts, which directly yields the expression for the mean kinetic energy:
\begin{equation*}
    \langle E_\mathrm{kin} \rangle = \frac{3}{2}k_BT.
\end{equation*}
Substituting to Eq.~\eqref{apx:eq:pgas}, we conclude that the thermal pressure in the case of the classical ideal gas is given by:
\begin{equation}
    \boxed{P_{\text{gas}} = \frac{2}{3}u = \frac{2}{3} n \langle E_{\text{kin}} \rangle = nk_BT = \frac{\rho}{\mu m_H}k_BT}
\end{equation}
which represents the well-known ideal gas law, emerging from considering non-relativistic classical particles. Here, $\mu$ is the mean molecular weight, $m_H$ is the mass of the hydrogen atom, and $k_B$ is the Boltzmann constant.
Notice that this law arises from the generalized equation of state with $\chi_\rho = \chi_T = 1$, highlighting the intrinsic connection between microscopic particle dynamics and macroscopic thermodynamic properties.



% \subsection{The photon gas}
\subsection{The ideal quantum gas}

According to statistical quantum physics, a collection of many indistinguishable particles in thermal equilibrium exhibits a momentum distribution given by:
    \begin{equation}
    f(\boldsymbol{p}) = \frac{1}{e^{(E(p) - \mu) / (k_B T)} \pm 1},
    \end{equation}
where the ``$\pm$'' sign takes the positive value for fermions and the negative value for bosons. In the first case, the distribution is referred to as the \textit{Fermi-Dirac distribution}, while in the second case, it is called the \textit{Bose-Einstein distribution}. The quantity $\mu$, known as the chemical potential, represents a form of potential energy that is related to the number density and temperature of the particles.

In the classical limit, that is, in the limit where the average distance between particles is much greater than the de Broglie wavelength of the particles, the above momentum distribution function can be approximated by the Maxwell-Boltzmann distribution. In the opposite case, quantum mechanical phenomena cannot be neglected. At this stage, it becomes essential to reference two fundamental principles of quantum mechanics. The first is the \textit{Heisenberg's uncertainty principle}:
\begin{equation}
\Delta x \Delta p_x \geq h \longrightarrow (\Delta x)^3 (\Delta p)^3 \geq h^3 \equiv V_x V_p \geq h^3,
\end{equation}
where $V_x$ is the geometric volume and $V_y$ is the volume in momentum space. An interpretation of this fundamental principle of quantum mechanics is that phase space is quantized, i.e., the minimum volume that a fermion can occupy is:
$$\Delta x \Delta y \Delta z \Delta p_x \Delta p_y \Delta p_z = h^3,$$
which is called a ``quantum cell'' and is defined in the 6-dimensional phase space.

The second principle of quantum mechanics relevant to our discussion is the \textit{Pauli's exclusion principle}. According to this principle, it is not possible for two or more fermions to exist with all their quantum numbers identical within a quantum cell. As a consequence of both the exclusion principle and the uncertainty principle, the number of energy states corresponding to momenta between $p$ and $p + dp$ is given by:
$$V_x V_p \geq h^3 \longrightarrow g(s) = g\frac{V_x V_p}{h^3},$$
where $g$ is the degeneracy of the energy states per quantum cell in the phase space and $V_x V_p = h^3$ is the minimum volume according to the uncertainty principle. The quantity $g(s)$ is called the ``density of states'' and for fermions, it is assumed that $g=2$ so that the density of states becomes:
$$ \frac{V_x}{h^3} 8 \pi p^2dp. $$
It follows that the number of fermions, $N(p)dp$, is given by:
$$N(p)dp \leq \frac{V_x}{h^3} 8 \pi p^2dp, $$
with equality holding when all the lowest possible energy states are occupied. Indeed, this occurs in a fully degenerate fermion gas, as we will discuss later, for which we can simply write:
$$N(p) = g \frac{V_x V_p}{h^3} = 2 \frac{V_x V_p}{h^3}.$$
Based on the above, and because $N = n V_x$, the number of fermions in a given phase space with momenta between $p$ and $p + dp$ is given by:
\begin{equation}
    n(p) dp = f g \frac{V_p}{h^3} = f g \frac{4\pi p^2 dp}{h^3} = \frac{g}{e^{(E - \mu)/k_B T} + 1} \frac{4\pi p^2}{h^3} dp.
\end{equation}

Suppose we have a distribution of electrons (fermions); then the number density of the electrons is given by the relation:
\begin{equation}
    n_e = \int_0^{\infty} n_e (p) dp =  \frac{8\pi}{h^3} \int_0^{\infty} \frac{p^2 dp}{e^{(E - \mu)/k_B T} + 1}.
\end{equation}
Under conditions of high pressure and low temperature, the thermal energy of an electron is much smaller than its rest energy. As a result, the electrons occupy the lowest possible energy levels. All energy levels up to a certain maximum are filled, while those above this level remain empty. The highest energy level that the quantum system reaches is called the \textit{Fermi energy}, $\epsilon_F$, which is defined as:
$$\mu = mc^2 + \epsilon_F.$$
Thus, when $k_B T \ll \mu$ (or equivalently, $k_B T \ll \epsilon_F$), the electron gas is considered \textit{fully degenerate}.
Instead of the term "Fermi energy," we can say that there exists a highest momentum value in a fully degenerate fermion gas, which we correspondingly call the \textit{Fermi momentum}, $p_F$. The Fermi momentum is related to the Fermi energy as follows: for a non-relativistic gas, $\epsilon_F = p_F^2 / 2m_e$, and for a relativistic gas, $\epsilon_F = c\,p_F$.

Under the conditions $E \ll \mu$ and $T = 0$, the number density of electrons is given by:
\begin{equation}
    \label{apx:electron_number_density_degenerate_gas}
    n_e = \frac{8\pi}{h^3} \int_0^{p_F} \frac{p^2 dp}{e^{(E - \mu)/k_B T} + 1} = \frac{8\pi}{h^3} \int_0^{p_F} p^2 dp = \frac{8\pi}{3h^3} p_F^3,
\end{equation}
as $e^{(E - \mu)/k_B T} \rightarrow 0$. The upper limit of integration for a fully degenerate gas is the Fermi momentum, since, by definition, electrons cannot possess momentum (or energy) greater than this limit. Therefore, we have:
\begin{equation}
    \label{apx:eq:fermi_momentum}
    p_F = h \left( \frac{3 n_e}{8 \pi} \right)^{1/3} \longrightarrow p_F \propto n_e^{1/3}.
\end{equation}

Using the information discussed so far, we can easily derive the equation of state for both non-relativistic and relativistic degenerate gases. In the non-relativistic case, where the electrons are not moving at relativistic speeds, it holds that $E_\text{kin} = p^2 / 2m$ and $P = \frac{2}{3}n \langle E_\text{kin} \rangle$. Considering degeneracy, we can write:
\begin{align*}
    \langle E_\text{kin} \rangle &= \frac{1}{N} \int_0^{\epsilon_F} E N(E) dE = \frac{1}{N} \int_0^{p_F} \frac{p^2}{2m} \frac{2}{h^3} V_x 4\pi p^2 dp \\\\
    &= \frac{4\pi V_x}{m N h^3} \int_0^{p_F} p^4 dp = \frac{4\pi V_x}{m N h^3} \frac{p_F^5}{5} = \frac{p_F^2}{2m} \frac{8\pi V_x}{5Nh^3} p_F^3 \\\\
    &= \frac{p_F^2}{2m} \frac{8\pi V_x}{5Nh^3} \frac{3 V_p}{4\pi} = \frac{p_F^2}{2m} \frac{2 V_x V_p}{Nh^3} \frac{3}{5} \Rightarrow \\\\
    &\Rightarrow \langle E_{\text{kin}} \rangle = \frac{3}{5} \frac{p_F^2}{2m} = \frac{3}{5} \epsilon_F,
\end{align*}
where we have used $V_p = \frac{4}{3} \pi p_F^3$ and $V_x V_p = N h^3/2$. 
The pressure exerted by the degenerate electron gas is then given by:
$$P_{\text{deg}} = \frac{2}{3} n_e \langle E_{\text{kin}} \rangle  = \frac{2}{3} \frac{8\pi}{3h^3} p_F^3 \frac{3}{5}\frac{p_F^2}{2m_e} = \frac{16\pi}{30m_e h^3}p_F^5.$$
However, the Fermi momentum can be written according to Eq.~\eqref{apx:eq:fermi_momentum} as:
$$p_F = \left( \frac{3}{8\pi} \right)^{1/3} h n_e^{1/3} = \left(\frac{3}{8\pi} \right)^{1/3} h \left(\frac{\rho}{\mu m_H} \right)^{1/3},$$
where $\rho$ is the mass density and $\mu$ refers to the mean molecular weight, not the chemical potential. Combining the last two equations, we obtain the pressure as:
\begin{equation}
    \label{apx:eq:degenerate_eos_nr}
    \boxed{P_{\text{deg}}^{\text{NR}} = \frac{16 \pi h^2}{30m_e} \left[ \frac{3}{8\pi m_H} \left(\frac{Z}{A} \right) \right]^{5/3} \rho^{5/3} \longrightarrow P_{\text{deg}}^{\text{NR}} = K_{\text{NR}} \left( \frac{Z}{A} \right)^{5/3} \rho^{5/3}}
\end{equation}
This is the equation of state that describes a \textit{non-relativistic, degenerate electron gas}. The term $K_{\text{NR}}$ is a constant that depends on the mass of the fermions (in our case, electrons) while for the mean molecular weight, it holds that $\frac{Z}{A} = \frac{1}{\mu}$.
Similarly, we can derive the equation of state for a relativistic, degenerate gas by using $P = \frac{1}{3}n \langle E \rangle$ and $E = p\,c$:
\begin{equation}
    \label{apx:eq:degenerate_eos_er}
    \boxed{P_{\text{deg}}^{\text{ER}} = \frac{1}{8} \left( \frac{3}{\pi} \right)^{1/3} \frac{c\,h}{m^{4/3}} \rho^{4/3} \longrightarrow P_{\text{deg}}^{\text{ER}} = K_{\text{ER}} \rho^{4/3}}
\end{equation}
where $K_{\text{ER}}$ is a constant.

Finally, it is important to note that while the pressure of a degenerate gas is independent of temperature (i.e., it is not of thermal nature), the condition for the degeneracy of a gas depends both on the density and the temperature. Indeed, the results presented above were derived from a quantum mechanical treatment of fermions at absolute zero ($T=0$). It is easy to show that for the case of a non-relativistic gas, the Fermi energy is:
$$\epsilon_F = \frac{3\hbar^2}{m} \left[\left( \frac{Z}{A} \right) \frac{\rho}{m_H} \right]^{2/3} \longrightarrow \epsilon_F \propto n_e^{2/3} \propto \rho_e^{2/3}.$$
This means that the condition for degeneracy\footnote{We could also define the \textit{Fermi temperature} as $\epsilon_F = k_B T_F$ and use it in the condition for degeneracy instead of the Fermi energy.} can be written as:
$$k_B T \ll \epsilon_F \Rightarrow \frac{T}{\rho^{2/3}} \ll D,$$
where $D$ is a constant. Thus, for sufficiently high temperatures, the degeneracy condition can be lifted, and the gas can transition back to a state akin to an ideal gas.

The case of a partially degenerate ideal gas, where $k_B T \sim \epsilon_F$, is significantly more complex when calculating the required physical quantities. In this scenario, not all energy levels are fully occupied, meaning the momentum distribution no longer has an upper limit at the Fermi momentum. Due to the increased complexity in modeling this case, we will not further examine this topic here.

\end{document}