\documentclass[main.tex]{subfiles}

\begin{document}
    \chapter*{Acknowledgements}
    First and foremost, I wish to thank my family for their unwavering support and encouragement throughout this journey. Their belief in me and their endless generosity in every imaginable way have been my foundation during these years.\\

    \noindent To my wife, I owe a deep debt of gratitude for her sacrifices and support, especially for accompanying me to Crete to pursue my PhD and for giving me the priceless gift of our beautiful daughter. To Katerina, my daughter, thank you for being my source of joy and keeping me sane through the toughest of times. \\

    \noindent I am incredibly grateful to my supervisor, John, whose support, understanding, and friendship have been indispensable. His guidance has not only steered my research but also shaped me as an academic and as a person.\\

    \noindent My heartfelt thanks go to Vassilis Charmandaris, the director of IA-FORTH, for his trust and support throughout my PhD. I am also deeply thankful to both Iossif Papadakis and Vassilis Charmandaris for offering me the opportunity to serve as a teaching assistant in their courses. These roles allowed me to sharpen my skills and gain invaluable experience, for which I am truly grateful. \\

    \noindent To Anna Steiakaki, Tasos Kougentakis, and Panagiotis Evangelopoulos, thank you for your kindness and support, which made my daily routine not just manageable but genuinely enjoyable. Your presence brightened even the busiest of days and brought much-needed moments of joy. \\

    \noindent To my fellow PhD students Elias Kyritsis and Nikos Mandarakas, thank you for your companionship, thought-provoking conversations, and for dedicating time from your personal lives to support my teaching initiatives. \\

    \noindent I would also like to express my heartfelt gratitude to my extended family---Christos Melidis, Miltos Papaminas, Telemachos Athanasiadis, Despoina Chasapi, and Despoina Konstantinou---for their constant presence and emotional support when I needed it most. Your friendship has transcended into something much deeper, and I am truly blessed to have you in my life. \\

    \noindent Last but not least, I would also like to thank the secretary of the Physics Department, Maria Matalliotaki and the secretary of IA-FORTH Eleftheria Tsentelierou, for their invaluable assistance in navigating the bureaucracy and handling the technical details of every aspect of my work. Your support and efficiency have made an enormous difference throughout this journey. \\

    \noindent Finally, to everyone who has played a part in this journey, no matter how big or small, thank you for helping make this achievement possible.\\

    \noindent The work presented in this thesis was supported by the Stavros Niarchos Foundation (SNF) and the Hellenic Foundation for Research and Innovation (H.F.R.I.) under the 2nd Call of ``Science and Society'' Action Always strive for excellence -- ``Theodoros Papazoglou'' (Project Number: 01431). I further acknowledge support from the European Commission under project ARGOS-CDS (Grant Agreement number: 101094354).
\end{document}